Equalizer Programming Guide

### 2008-10-07 Developer/ProgrammingGuide/paper.tex
Programming Guide: update for 2D pixel compounds
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

\date{
  \vspace{2cm}
  \begin{figure}[ht]
    \hspace{-4.5cm}\includegraphics[width=22cm]{images/teaser.pdf}
  \end{figure}
  \vspace{1cm}
%  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
  Version 1.3.9 for Equalizer 0.5.4+svn\\[\medskipamount]
  \today
}

@@

\clearpage
\thispagestyle{empty}
\tableofcontents
\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.2     & Apr 15, 2008 & Version for Equalizer 0.5\\
    1.3.1   & May 02, 2008 & Add \sref{sStatistics}, improved images based on 1.2 print\\
    1.3.2   & May 05, 2008 & Add network distribution on eqPly model data\\
    1.3.3   & May 10-19, 2008 & add \aref{aFileFormat}\\
    1.3.4   & July 3, 2008 & Update \sref{sStatistics}\\
    1.3.5   & July 7, 2008 & Update after namespace naming cleanup\\
    1.3.6   & July 20, 2008 & Add accum and samples to \aref{aFileFormat}, add \sref{sDrawableConfig}\\
    1.3.7   & Aug 22, 2008 & Update \sref{sEqplyWIndow} with OSWindow changes\\
    1.3.8   & Aug 25, 2008 & Update \sref{sEventHandling} with OSWindow changes\\
    1.3.9   & Aug 29, 2008 & Update \sref{sEventHandling}: event handling cleanup\\
    \hline
  \end{tabularx}}
\thispagestyle{empty}
\clearpage

@@

2D decompositions can be used by all types of applications, but should
be combined with DB compounds to reduce the data per node, if possible.

@@

@@

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/Pixel.pdf}
  {\caption{\small A pixel compound}}
\end{wrapfigure}
Pixel compounds are similar to 2D compounds. The frustra of the
source rendering units are modified so that each unit renders an evenly
distributed subset of pixels.

@@

Pixel compounds work well for purely fill-limited applications,
techniques like frustum culling do not reduce the rendered data for the
source rendering resources.

OpenGL functionality influenced by the raster position will not work
correctly with pixel compounds, or needs at least special
attention. Among them are: lines, points, sprites, glDrawPixels,
glBitmap, glPolygonStipple.

Pixel compounds are ideal for raytracing, which is highly fill-limited
and needs the full database for rendering anyway. Volume rendering
applications should choose this mode over 2D compounds.

@@

Some applications might rely on a shared filesystem to access the data,
for example when out-of-core algorithms are used. Other applications
prefer to load the data only on the application process, and use
distributed objects to synchronize the scene data with the render
clients.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/modelDist.pdf}
  {\caption{\small\label{fModelDist}Scene Data in eqPly}}
\end{wrapfigure}
The identifer of the root object of this distributed hierarchy is passed
as part of the \textsf{InitData}, and used during remote node
initialization to map the model. \fig{fModelDist} shows the UML
hierarchy of the model and distribution classes.

Each \textsf{VertexBufferDist} object has one node of the model's data
tree. It is serializing the data for this node. Furthermore, it mirrors
the kD-tree by having a \textsf{VertexBufferDist} child for each child
of its corresponding tree node. During serialization, the identifier of
these children is sent to the remote nodes, which reconstruct the mirror
distribution hierarchy and model data tree based on this data.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/objects.pdf}
  {\caption{\small\label{fObjects}Scene Graph Distribution}}
\end{wrapfigure}
Applications distributing a dynamic scene graph use the frame data
instead of the init data as the entry point to their scene graph data
structure. \fig{fObjects} shows one possible implementation, where the
identifier and version of the scene graph root are transported using the
frame data. The scene graph root then serializes and de-serializes his
immediate children by transferring their identifier and current version,
similar to the static distribution done by \textsf{eqPly}.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/osWindow.pdf}
  {\caption{\small\label{fOSWindow}eq::OSWindow UML class hierarchy}}
\end{wrapfigure}
All window-system specific functionality is implemented by a
specialization of \textsf{eq::OS\-Window}. The \textsf{OSWindow} class
defines the minimal interface to be implemented for a new window
system. Each \textsf{Window} uses one \textsf{OSWindow} during
execution. This separation allows an easy implementation and adaption to
another window system or application.

@@

The initialization sequence uses multiple, overrideable task
methods. The main task method \textsf{configInit} calls first
\textsf{configInitOSWindow}, which creates and initializes the
\textsf{OSWindow} for this window. The \textsf{OSWindow} initialization
code is implementation specific. If the \textsf{OSWindow} was
initialized successfully, \textsf{configInit} calls
\textsf{configInitGL}, which performs the generic OpenGL state
initialization. The default implementation sets up some typical OpenGL
state, e.g., it enables the depth test.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/DB.png}
  {\caption{\small\label{fDB}Destination view of a DB compound}}
\end{wrapfigure}
After the basic view setup, a directional light is configured, and the
model is positioned using the camera parameters from the frame data. The
camera parameters are transported using the the frame data to ensure
that all channels render a given frame using the same position.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/render.pdf}
  {\caption{\small\label{fRender}Main Render Loop}}
\end{wrapfigure}
The model data is spatially organized in an 3-dimensional
kD-tree\footnote{See also \link{http://en.wikipedia.org/wiki/Kd-tree}}
for efficient view frustum culling. When the model is loaded by
\textsf{Node::config\-Init}, it is preprocessed into the kD-tree and each
node of the tree gets a database range assigned. The root node
has the range [0, 1], its left child [0, 0.5] and its right child [0.5,
1], and so on for all nodes in the tree. The preprocessed model is saved
in a binary format for accelerating subsequent use.

The rendering loop maintains a list of candidates to render, which
initially contains the root node. Each candidate of this list is tested
for full visiblity against the frustum and range, and rendered if
visible. It is dropped if it is fully invisible or fully out of
range. If it is partially visible or partially in range, the children of
the node are added to the candidate list. \fig{fRender} shows a flow
chart of the rendering algorithm, which performs efficient view frustum
and range culling.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/eventFilter.pdf}
  {\caption{\small\label{fEventProcessing}Event Processing}}
\end{wrapfigure}
The purpose of the \textsf{OSWindow} processing method,
\textsf{processEvent}, is to perform window-system specific event
processing. For example, \textsf{AGLWindow::processEvent} calls
\textsf{aglUpdateContext} whenever a resize event is received. For
further processing, the \textsf{Event} is passed on to
\textsf{Window::processEvent}. This \textsf{Event} no longer contains
the native event.

\textsf{Window::processEvent} is responsible for either handling
the event locally, or for translating it into a generic
\textsf{ConfigEvent}. The config events are send to the application
thread using \textsf{Config::sendEvent}.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/eventUML.pdf}
  {\caption{\small\label{fEventUML}UML Diagram for Event Handling Classes}}
\end{wrapfigure}
Currently, Equalizer provides an \textsf{AGLEvent\-Handler},
\textsf{GLXEventHandler} and \textsf{WGL\-Event\-Handler}, which handle
events for an \textsf{AGLWindowIF}, \textsf{GLXWindowIF} and
\textsf{WGLWindowIF}, respectively. \fig{fEventUML} illustrates the
class hierarchy for event processing.

The \textsf{OSWindow} implementation is responsible for setting up and
de-initializing event handling.

@@

\subsubsection{Parallel Direct Send Compositing}

>>>>

\date{
  \vspace{2cm}
  \begin{figure}[ht]
    \hspace{-4.5cm}\includegraphics[width=22cm]{images/teaser.pdf}
  \end{figure}
  \vspace{1cm}
%  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
  Version 1.3.10 for Equalizer 0.5.5\\[\medskipamount]
  \today
}

@@

\clearpage
\thispagestyle{empty}
\tableofcontents
\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.2     & Apr 15, 2008 & Version for Equalizer 0.5\\
    1.3.1   & May 02, 2008 & Add \sref{sStatistics}, improved images based on 1.2 print\\
    1.3.2   & May 05, 2008 & Add network distribution on eqPly model data\\
    1.3.3   & May 10-19, 2008 & add \aref{aFileFormat}\\
    1.3.4   & July 3, 2008 & Update \sref{sStatistics}\\
    1.3.5   & July 7, 2008 & Update after namespace naming cleanup\\
    1.3.6   & July 20, 2008 & Add accum and samples to \aref{aFileFormat}, add \sref{sDrawableConfig}\\
    1.3.7   & Aug 22, 2008 & Update \sref{sEqplyWIndow} with OSWindow changes\\
    1.3.8   & Aug 25, 2008 & Update \sref{sEventHandling} with OSWindow changes\\
    1.3.9   & Aug 29, 2008 & Update \sref{sEventHandling}: event handling cleanup\\
    1.3.10  & Oct 7, 2008  & Updated to reflect 2D pixel compounds\\
    \hline
  \end{tabularx}}
\thispagestyle{empty}
\clearpage

@@

2D decompositions can be used by all types of applications, but should
be combined with DB compounds to reduce the data per node, if
possible. In most cases, a \textsf{loadBalancer} should be used to
automatically adapt the tiling each frame, based on the current
rendering load.

2D compounds in Equalizer are configured using the \textsf{viewport}
parameter, using the values \textsf{[ x y width height]} in normalized
coordinates. The viewport defines the area of the parent (destination)
channel to be rendered for each child. Each child compound uses an
output frame, which is connected to an input frame on the destination
channel. The destination channel can also be used as a source channel,
in which case it renders in place and no output frame is needed.

@@

DB compounds in Equalizer are configured using the \textsf{range}
parameter, using the values \textsf{[ begin end ]} in normalized
coordinates. The range defines the start and end point of the
application's database to be rendered. The value has to be interpreted
by the application's rendering code accordingly.  Each child compound
uses an output frame, which is connected to an input frame on the
destination channel. For more than two contributing channels, it is
recommended to configure parallel direct send compositing, as described
in \sref{sDirectSend}.

@@

Eye compounds in Equalizer are configured using the \textsf{eye}
parameter, limiting the child to render the \textsf{[ LEFT ]} or
\textsf{[ RIGHT ]} eye. Each child compound uses an output frame, which
is connected to an input frame on the destination channel. The
destination channel can also be used to render an eye pass, in which case
it renders in the correct stereo buffer and no output frame is needed.

@@

Pixel compounds are similar to 2D compounds. The frustra of the
source rendering units are modified so that each unit renders an evenly
distributed subset of pixels.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/Pixel.pdf}
  {\caption{\small A pixel compound}}
\end{wrapfigure}
Pixel compounds work well for purely fill-limited applications.
Techniques like frustum culling do not reduce the rendered data for the
source rendering resources.

OpenGL functionality influenced by the raster position will not work
correctly with pixel compounds, or needs at least special
attention. Among them are: lines, points, sprites, glDrawPixels,
glBitmap, glPolygonStipple. The application can query the current pixel
parameters at run-time to adjust the rendering accordingly.

Pixel compounds are ideal for raytracing, which is highly fill-limited
and needs the full database for rendering anyway. Volume rendering
applications are also well suited for this mode, and should choose it
over 2D compounds.

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/pixelKernel.pdf}
  {\caption{\label{fPixelKernel}\small Pixel compound kernel}}
\end{wrapfigure}
Pixel compounds in Equalizer are configured using the \textsf{pixel}
parameter, using the values \textsf{[ x y width height]}  to configure
the size and offset of the sampling kernel. The width and height of the
sampling kernel define how many pixels are skipped in the x and y
direction, respectively. The x and y offset define the index of the
source channel within the kernel. They have to be smaller than the size
of the kernel. \fig{fPixelKernel} illustrates these parameters, and
\fig{fPixelKernels} shows some example kernels for a four-to-one pixel decomposition.

The destination channel can also be used as a source channel. Contrary
to the other compound modes, it also has to use an output and
corresponding input frame. During rendering, the frustum is 'squeezed'
to configure the pixel decomposition. The destination channel can
therefore not be rendered in place, like with the other compound modes.
\begin{figure}[ht!]\center
  \includegraphics[width=.618\textwidth]{images/pixelKernels.pdf}
  {\caption{\small\label{fPixelKernels}Example pixel kernels for a
      four-to-one pixel compound}}
\end{figure}

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \vspace{-2em}\includegraphics[width=.382\textwidth]{images/modelDist.pdf}
  {\caption{\small\label{fModelDist}Scene Data in eqPly}}
  \vspace{-2em}
\end{wrapfigure}
Some applications might rely on a shared filesystem to access the data,
for example when out-of-core algorithms are used. Other applications
prefer to load the data only on the application process, and use
distributed objects to synchronize the scene data with the render
clients.

@@

The identifer of the root object of this distributed hierarchy is passed
as part of the \textsf{InitData}, and used during remote node
initialization to map the model. \fig{fModelDist} shows the UML
hierarchy of the model and distribution classes.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/objects.pdf}
  {\caption{\small\label{fObjects}Scene Graph Distribution}}
\end{wrapfigure}
Each \textsf{VertexBufferDist} object has one node of the model's data
tree. It is serializing the data for this node. Furthermore, it mirrors
the kD-tree by having a \textsf{VertexBufferDist} child for each child
of its corresponding tree node. During serialization, the identifier of
these children is sent to the remote nodes, which reconstruct the mirror
distribution hierarchy and model data tree based on this data.

@@

Applications distributing a dynamic scene graph use the frame data
instead of the init data as the entry point to their scene graph data
structure. \fig{fObjects} shows one possible implementation, where the
identifier and version of the scene graph root are transported using the
frame data. The scene graph root then serializes and de-serializes his
immediate children by transferring their identifier and current version,
similar to the static distribution done by \textsf{eqPly}.

@@

All window-system specific functionality is implemented by a
specialization of \textsf{eq::OS\-Window}. The \textsf{OSWindow} class
defines the minimal interface to be implemented for a new window
system. Each \textsf{Window} uses one \textsf{OSWindow} during
execution. This separation allows an easy implementation and adaption to
another window system or application.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/osWindow.pdf}
  {\caption{\small\label{fOSWindow}eq::OSWindow UML class hierarchy}}
\end{wrapfigure}
The initialization sequence uses multiple, overrideable task
methods. The main task method \textsf{configInit} calls first
\textsf{configInitOSWindow}, which creates and initializes the
\textsf{OSWindow} for this window. The \textsf{OSWindow} initialization
code is implementation specific. If the \textsf{OSWindow} was
initialized successfully, \textsf{configInit} calls
\textsf{configInitGL}, which performs the generic OpenGL state
initialization. The default implementation sets up some typical OpenGL
state, e.g., it enables the depth test.

@@

After the basic view setup, a directional light is configured, and the
model is positioned using the camera parameters from the frame data. The
camera parameters are transported using the the frame data to ensure
that all channels render a given frame using the same position.

@@

The model data is spatially organized in an 3-dimensional
kD-tree\footnote{See also \link{http://en.wikipedia.org/wiki/Kd-tree}}
for efficient view frustum culling. When the model is loaded by
\textsf{Node::config\-Init}, it is preprocessed into the kD-tree and each
node of the tree gets a database range assigned. The root node
has the range [0, 1], its left child [0, 0.5] and its right child [0.5,
1], and so on for all nodes in the tree. The preprocessed model is saved
in a binary format for accelerating subsequent use.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/render.pdf}
  {\caption{\small\label{fRender}Main Render Loop}}
\end{wrapfigure}
The rendering loop maintains a list of candidates to render, which
initially contains the root node. Each candidate of this list is tested
for full visiblity against the frustum and range, and rendered if
visible. It is dropped if it is fully invisible or fully out of
range. If it is partially visible or partially in range, the children of
the node are added to the candidate list. \fig{fRender} shows a flow
chart of the rendering algorithm, which performs efficient view frustum
and range culling.

@@

The purpose of the \textsf{OSWindow} processing method,
\textsf{processEvent}, is to perform window-system specific event
processing. For example, \textsf{AGLWindow::processEvent} calls
\textsf{aglUpdateContext} whenever a resize event is received. For
further processing, the \textsf{Event} is passed on to
\textsf{Window::processEvent}. This \textsf{Event} no longer contains
the native event.

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/eventFilter.pdf}
  {\caption{\small\label{fEventProcessing}Event Processing}}
\end{wrapfigure}
\textsf{Window::processEvent} is responsible for either handling
the event locally, or for translating it into a generic
\textsf{ConfigEvent}. The config events are send to the application
thread using \textsf{Config::sendEvent}.

@@

Currently, Equalizer provides an \textsf{AGLEvent\-Handler},
\textsf{GLXEventHandler} and \textsf{WGL\-Event\-Handler}, which handle
events for an \textsf{AGLWindowIF}, \textsf{GLXWindowIF} and
\textsf{WGLWindowIF}, respectively. \fig{fEventUML} illustrates the
class hierarchy for event processing.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/eventUML.pdf}
  {\caption{\small\label{fEventUML}UML Diagram for Event Handling Classes}}
\end{wrapfigure}
The \textsf{OSWindow} implementation is responsible for setting up and
de-initializing event handling.

@@

\subsubsection{\label{sDirectSend}Parallel Direct Send Compositing}
